Tidy data structure to support exploration and modeling of temporal-context data

### 2018-09-03 main.Rmd
fixed typos in abstract and intro

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Mining temporal-context data for information is often inhibited by a multitude of time formats: irregular or multiple time intervals, multiple observational units or repeated measurements on multiple individuals, heterogeneous data types, nested and crossed factors indicating hierarchical sub-groups. Time series models, in particular, the software supporting time series forecasting makes strict assumptions on data that needs to be provided, typically a matrix of numeric data with an implicit time index. Going from raw data to model-ready data is painful. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and forecasting routines. Tidy data principles are applied, and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) including an explicitly declared index variable representing time, (3) incorporating a "key" comprised of single or multiple variables to uniquely identify units over time, using a syntax-based and user-oriented approach in which it imposes nested or crossed structures on the data. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound data pipeline facilitates a fluent and transparent workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline structure and usage. The infrastructure of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal data, time series, data structures, data wrangling, tidy data, R, forecasting, data science, exploratory data analysis"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: [references.bib, rpkgs.bib]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

Temporal-context data, $X_{jt}$, consist of $j = 1, \dots, N_{i}$ observational units indexed at different time points, $1 \le t \le T$. Time forms the contextual basis of temporal data, but it can arrive in many possible formats, which makes preparing it for modeling and forecasting can be painful. For example, data can be recorded at various time resolutions (hours, minutes, and seconds), and they are typically associated with different time zones with adjustments like summer time. It could be irregularly recorded, which is particularly true with longitudinal measurements like patient visits to a doctor's office. Temporal data also often contains with rich information other than the time: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors involving nested or crossed structures.

>>>>

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Mining temporal-context data for information is often inhibited by a multitude of time formats: irregular or multiple time intervals, multiple observational units or repeated measurements on multiple individuals, heterogeneous data types, nested and crossed factors indicating hierarchical sub-groups. Time series models, in particular, the software supporting time series forecasting makes strict assumptions on data to be provided, typically a matrix of numeric data with an implicit time index. Going from raw data to model-ready data is painful. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and forecasting routines. Tidy data principles are applied, and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) including an explicitly declared index variable representing time, (3) incorporating a "key" comprised of single or multiple variables to uniquely identify units over time, using a syntax-based and user-oriented approach in which it imposes nested or crossed structures on the data. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound data pipeline facilitates a fluent and transparent workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline structure and usage. The infrastructure of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal data, time series, data structures, data wrangling, tidy data, R, forecasting, data science, exploratory data analysis"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: [references.bib, rpkgs.bib]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

Temporal-context data, $X_{jt}$, consist of $j = 1, \dots, N_{i}$ observational units indexed at different time points, $1 \le t \le T$. Time forms the contextual basis of temporal data, but it can arrive in many possible formats, which makes preparing it for modeling and forecasting painful. For example, data can be recorded at various time resolutions (hours, minutes, and seconds), and they are typically associated with different time zones with adjustments like summer time. It could be irregularly recorded, which is particularly true with longitudinal measurements like patient visits to a doctor's office. Temporal data also often contains rich information in addition to time: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors involving nested or crossed structures. The data problems are grouped into two types of analysis, time series and longitudinal.
